\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2023-03-08 main.tex

@@

\ignore{
\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}
}

@@

Dynamically typed programming languages are popular in education and the software industry. While presenting a low barrier to entry, they suffer from run-time type errors and longer-term problems in code quality and maintainability. Statically typed languages, while showing strength in these aspects, lack in learnability and ease of use. In particular, fixing type errors poses challenges to both novice users and experts. Further, compiler type error messages are presented in a static way that is biased toward the first occurrence of the error in the program code. To help users resolve such type errors we introduce \chameleon{}, a type debugging tool that presents type errors to the user in an unbiased way, allowing them to explore the full context of where the errors could occur. Programmers can interactively verify the steps of reasoning against their intention. Through three studies involving real programmers, we showed that  \chameleon{} is more effective in fixing type errors than traditional text-based error messages. This difference is more significant in harder tasks. Further, programmers actively using \chameleon{}'s interactive features are shown to be more efficient in fixing type errors than passively reading the type error output.
\end{abstract}

>>>>

\makeatletter % changes the catcode of @ to 11
\newcommand{\linebreakand}{%
  \end{@IEEEauthorhalign}
  \hfill\mbox{}\par
  \mbox{}\hfill\begin{@IEEEauthorhalign}
}
\makeatother % changes the catcode of @ back to 12

@@

%\ignore{
\author{\IEEEauthorblockN{Shuai Fu}
\IEEEauthorblockA{
\textit{Faculty of Information Technology} \\
\textit{Monash University}\\
Clayton, Australiay \\
\texttt{shuai.fu@monash.edu}}
\and
\IEEEauthorblockN{Tim Dwyer}
\IEEEauthorblockA{
\textit{Faculty of Information Technology} \\
\textit{Monash University}\\
Clayton, Australiay \\
\texttt{tim.dwyer@monash.edu}}
\and
\IEEEauthorblockN{Peter J. Stuckey}
\IEEEauthorblockA{
\textit{Faculty of Information Technology} \\
%\textit{Data Science and Artificial Intelligence} \\
\textit{Monash University}\\
Clayton, Australia \\
\texttt{peter.stuckey@monash.edu}}
\and
\linebreakand 
\IEEEauthorblockN{Jackson Wain}
\IEEEauthorblockA{\textit{Faculty of Information Technology} \\
\textit{Monash University}\\
Clayton, Australia \\
\texttt{0000-0003-2006-3538}
}
\and
\IEEEauthorblockN{Jesse Linossier}
\IEEEauthorblockA{\textit{Faculty of Information Technology} \\
\textit{Monash University}\\
Clayton, Australia \\
\texttt{0000-0001-6782-7019}
}

}
%}

@@

Dynamically typed programming languages are popular in education and the software industry. While presenting a low barrier to entry, they suffer from run-time type errors and longer-term problems in code quality and maintainability. Statically typed languages, while showing strength in these aspects, lack in learnability and ease of use. In particular, fixing type errors poses challenges to both novice users and experts. Further, compiler type error messages are presented in a static way that is biased toward the first occurrence of the error in the program code. To help users resolve such type errors we introduce \chameleon{}, a type debugging tool that presents type errors to the user in an unbiased way, allowing them to explore the full context of where the errors could occur. Programmers can interactively verify the steps of reasoning against their intention. Through three studies involving actual programmers, we showed that \chameleon{} is more effective in fixing type errors than traditional text-based error messages. This difference is more significant in harder tasks. Further, programmers actively using \chameleon{}'s interactive features are shown to be more efficient in fixing type errors than passively reading the type error output.
\end{abstract}

### 2023-03-08 sections/chameleon.tex

The \chameleon{} debugging interface provides three main features to visualize the relation between the error statement and locations in code and to explore the explanation of each error location given by the type inference engine. %Further, programmers  can turn on and off features to suit their preferences and debugging needs.

@@

Using only constraints from the MUS is sufficient to locate the type error, but to recover types from type errors we need  constraints from parts of the program that are irrelevant to the type error.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are theoretically correct, they introduce a new type variable \texttt{a}, making the error message harder to understand. To solve this issue in \chameleon{}, for each constraint \texttt{c} in the MUS, we find a maximally satisfiable subset (MSS) from all the constraints that contain every other element of MUS but not \texttt{c}. These maximally satisfiable subsets, while not helpful to inform error locations, will produce the most concrete conflicting types, see Fig.~\ref{fig:compare-to-original}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 6mm 0mm 0mm]{images/compare-to-original.pdf}
    \caption{
Reporting the same type error, Chameleon uses
\texttt{Int -> a} and \texttt{Char -> a}, while \chameleon{} uses the 
concrete types \texttt{Int -> Bool} and \texttt{Char -> Bool}.
    }
    \label{fig:compare-to-original}
\end{figure}

Using concrete types provides extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, a type of \texttt{(a, Float)} does not preserve such information. If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.

@@

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system  in ~\cite{jun_explaining_2002},  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by annotating nodes in the abstract syntax tree with constraints and the typing rules used. We generate an inference history from constraints and accompanying annotations.

>>>>

The \chameleon{} debugging interface provides three main features to visualize and explain type errors.

% the relation between the error statement and locations in code and to explore the explanation of each error location given by the type inference engine. 
%Further, programmers  can turn on and off features to suit their preferences and debugging needs.

@@

Using only constraints from the MUS is sufficient to locate the type error, but to recover types from type errors we need constraints from parts of the program that are irrelevant to the type error.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are theoretically correct, they introduce the notation \texttt{a}, which denotes a generic type variable that can be any type, making the error message harder to understand. To solve this issue in \chameleon{}, for each constraint \texttt{c} in the MUS, we find a maximally satisfiable subset (MSS) from all the constraints that contain every other element of MUS but not \texttt{c}. These maximally satisfiable subsets, while not helpful in error localization, will produce the most concrete types, see Fig.~\ref{fig:compare-to-original}. Using concrete types, such as \texttt{Int} and \texttt{String},  provides extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, a type of \texttt{(a, Float)} does not preserve such information.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 6mm 0mm 0mm]{images/compare-to-original.pdf}
    \caption{
Reporting the same type error, Chameleon uses more abstract types
\texttt{Int -> a} and \texttt{Char -> a}, while \chameleon{} uses the 
concrete types (types that do not contain type variables) \texttt{Int -> Bool} and \texttt{Char -> Bool}.
    }
    \label{fig:compare-to-original}
\end{figure}

% If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.

@@

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system  in ~\cite{jun_explaining_2002},  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by annotating nodes in the abstract syntax tree with constraints and the type inference rules used. We generate an inference history from constraints and accompanying annotations.

### 2023-03-08 sections/discussion.tex

\subsection{Forming Debugging Plans}
From the results of Study 2, we found that programmers who use the interactive tool fix type errors faster than the ones who passively read the error output. This effect is stronger in harder tasks. We speculate that one factor of this result is that  \chameleon{} helps to develop debugging plans. We observed that when working with \chameleon{}, programmers form different debugging tactics to attack the problem. Among the \textit{high} interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some programmers navigate to both ends of the deduction steps where types are normally grounded and concrete. In contrast, \textit{minimal} interactive participants generally form similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.

@@

>>>>

\subsection{Forming Debugging Plans}
From the results of Study 2, we found that programmers who use the interactive tool fix type errors faster than the ones who passively read the error output. This effect is stronger in harder tasks. We speculate that one factor of this result is that  \chameleon{} helps to develop debugging plans. We observed that when working with \chameleon{}, programmers form different debugging plans to attack the problem. Among the \textit{high} interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some navigate to both ends of the deduction chain where types are normally grounded and concrete. In contrast, \textit{minimal} interactive participants generally form similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.

@@

% \subsection{Implication for future studies}
% Our study aims to contribute to the wider understanding of program comprehension in the context of advanced type systems. One promising research area to extend our results is to apply the design of \chameleon{} to other languages with high demand for explainability\cite{ferdowsi_usability_2023}.

% Colorization and overlay widgets are used to visualize relations and orders that may not conform to the source code structures. These techniques fit well to types because the introduction and elimination of typing constraints tend to be scattered throughout the program. A interesting question is whether these techniques can also be applied to other program properties, such as memory allocation and freeing, as well as message dispatching and consumption, which also exhibit similar patterns.

% Finally, we employed MUS (minimal unsatisfiable subset) for fault localization and types recovery, but we are also curious about the possibility of further analysis based on MUS to power advanced debugging systems, such as efficient and accurate type error repair by combining our results with mutation testing or solution searching based on machine learning.
